A continental-scale analysis finds generalists, with large clutch size and residual brain mass, are the most urban-tolerant of birds

### 2018-05-20 ms.Rmd
some more tweaks and changes based on RM comments

---
title: "A continuous urbanization index to assess phylogenetic adaptations in urban bird living"

@@

# ABSTRACT
Understanding the suite of ecological and life history traits which influence a species' tolerance to urbanization enables a predictive ability to elucidate those species which are most likely to persist in our increasingly urbanizing world. Previous work investigating these relationships has found contrasting patterns for a trait's association with urbanization. In a continental-scale analysis, incorporating 477 species and > 5,000,000 bird observations, we provide a novel and scalable methodology which assesses the ecological and life history traits which most influence a species' adaptability to persist in urban environments. We elucidate generalized, phylogenetically controlled, patterns: bird species which are specialists, with high clutch size, and large residual brain size are among the most urban-tolerant bird species. Conversely, specialized feeding strategies (i.e., insectivores and granivores) are negatively associated with urbanization. Our results suggest that the enhancement and persistence of avian biodiversity in urban environments relies on maintaining and developing a suite of habitats to serve a range of life history strategies.

@@

In order to best predict species' responses to future urbanization changes, it is essential to understand what ecological and life history traits are most associated with birds that are either successful or unsuccessful in urban environments [e.g., @mcclure1989characterizes; @bonier2007urban; @moller2009successful;  @mcdonnell2015adaptation]. To date, a wide array of traits have been used to predict presence of birds in urban environments. For instance, degree of sociality [@mcclure1989characterizes; @coleman1994neophobia; @jokimaki1998distribution], nesting substrate [@kark2007living; @croci2008does; @conole2011functional], diet [@beissinger1982effects; @fuller2008garden; @major2010museum; @evans2011makes], distribution size [@jokimaki2000artificial; @chace2006urban; @croci2008does], migratory status [@friesen1995effects], niche breadth [@clergeau2006avifauna; @kark2007living; @evans2011makes], and fecundity [@croci2008does; @moller2009successful; @vaugoyeau2016interspecific], among other traits (see Appendix 1), have been hypothesized to influence a species' presence in urban environments. Rather than a specific trait that predicts a species' ability to be found in urban environments, it is most likely a combination of traits. For instance, the presence of urban birds in Jerusalem was predicted by traits including diet, degree of sociality, sedentariness, and preferred nesting sites [@kark2007living], while a regional analysis throughout France and Switzerland demonstrated that urban birds preferred forested habitats, were sedentary, omnivorous, high-elevation nesters, and widely distributed [@croci2008does]. Notably, a wide range of results using the aforementioned traits have been reported in the literature, often contrasting with one another. Indeed, residual brain size was found to be important by Moller and Erritze [-@moller2015brain] and Maklakov *et al.* [-@maklakov2011brains] but Evans *et al.* [-@evans2011makes] and Kark *et al.* [-@kark2007living] found that brain size had no effect on species' response to urbanization. Similarly, contrasting results have been reported for annual fecundity [cf., @evans2011makes with @moller2009successful & @croci2008does] and niche breadth [cf., @kark2007living with @evans2011makes]. Forming generalizable patterns, at multiple spatial scales [@suarez2018scale], which associate ecological and life history traits with urbanization in birds is critical to help curb the impact urbanization has on avian biodiversity [@sol2017urbanisation].

The majority of previous studies place species in discrete groups such as urban avoiders, adapters, or exploiters [e.g., @mckinney2002urbanization; @mckinney2006urbanization; @kark2007living; @mcdonnell2015adaptation] or even binary grouping based on presence/absence within urban environments [e.g., @bonier2007urban; @moller2009successful]. These types of discrete response variables severely limits the capacity to generalize results as it assumes species within a group respond to urbanization equally [@evans2011makes]. Moreover, the terminology (i.e., urban adapters, avoiders, and exploiters) is used differently among studies [cf., [@kark2007living] and [@croci2008does]], potentially affecting interpretation between scientists and the public [@mcdonnell2015adaptation; @fischer2015categorizing]. In a notable exception, Evans *et al.* [-@evans2011makes] advocates for a continuous measure of urbanization in which a species is assessed on its entire distribution in response to urban environments.

@@

The aim of this study is to determine whether there is a suite of life history traits that predicts avian tolerance to urbanization at a continental scale. For such an analysis to have generality it must (1) be comprehensive in faunal coverage, (2) account for phylogenetic constraints, (3), include a broad-range of traits, and (4) employ a continuous measure of a specie's response to urbanization. We conduct an analysis that meets these criteria, using a series of publicly available datasets to implement a novel continuous urbanization index in order to assess ecological and life history traits which most influence a bird's response to urbanization. We hypothesized that a distinct suite of particular traits would be positively associated with urbanization (i.e., urbanization selects for) while a separate suite of traits would be negatively associated with urbanization (i.e., urbanization selects against). Our first objective was to investigate the presence of a phylogenetic signal in our urbanization index, as previous studies have found a lack of phylogenetic signal in urbanization [@evans2011makes]. Our second objective was to develop models which were and were not phylogenetically controlled to assess species' traits and their association with urbanization, using a suite of continuous and categorical predictor variables.

@@

We weighted the linear regression models [@solon2015weighting] using a source of error associated with our urbanization indices. This error was the  proportion of the number of total observations of a species divided by the number of unique localities a species was observed from. This gave more weight in the model-fitting procedure to those species that had more observations in more unique locations. This was done based on the assumption that those species which had the most observations would have the most reliable urbanization indices. This approach also addresses biases in sampling of eBird data in which some birds can have a high number of observations, but from known roosting sites [e.g., @callaghanavian].

@@

# DISCUSSION
Niche breadth, as measured by different types of generalism, was a significantly important life-history characteristic associated with the most urban-tolerant species in our analyses. This result was robust across non-phylogenetic and phylogenetic models, model-averaging approaches, and across phylogenetic uncertainty (Fig. S3). Conversely, specialized feeding strategies were negatively associated with urbanization: granivorous and insectivorous specialists were much less likely to be found in urban environments. These results suggest that urbanization is, in general, excluding specialist species [@sorace2009distribution]. As such, future urban development aimed at increasing avian diversity in urban areas should focus on maintaining a high diversity of habitat [@melles2003urban] both for foraging [@shochat2004linking], and nesting [@fernandez2001habitat], with a focus on intact vegetation [@sushinsky2013should] connected by corridors [@litteral2017role]. A species-specific approach to assess particular resources necessary for specialist species may be necessary [@fernandez2004spatial]. Furthermore, a focal point of avian biodiversity conservation should be on increasing insect abundance and diversity in urban environments [@baldock2015uk], which will in turn benefit avian biodiversity in urban environments.

The non-random distribution of the urbanization index across the phylogeny (Table 1) is consistent with the observation of reduced phylogenetic diversity in urban areas [@la2018phylogenetic], suggesting that particular phylogenetic constraints allow certain species to remain tolerant of urban environments [@morelli2016evidence; @ibanez2017global; @sol2017urbanisation]. This finding of a strong phylogenetic signal contrasts with previous studies [@evans2011makes], but this is likely a result of increased statistical power from our larger sample size: 477 species in this study compared with 55 species in Evans *et al.* [-@evans2011makes]. It's not surprising that a larger sample size finds a non-random pattern in urbanization across the bird phylogeny and we demonstrated the importance of treating species as non-independent by considering phylogeny in the context of the linear models. For example, some predictor variables became less crucial in the phylogenetic models compared with the non-phylogenetic models, including some feeding types (i.e., insect specialists) and habitat preferences (i.e., agricultural). For these variables, it is likely that there may be particular clades which include a number of species that are associated with both that particular trait and urbanization [see @felsenstein1985confidence]. Conversely, some of the most important predictor variables (e.g., clutch size and feeding habitat generalism; Figures 4 and 5) showed a stronger relationship (i.e., increased parameter estimates) with urbanization when phylogeny was accounted for, and this affect was robust relative to a consideration of phylogenetic uncertainty (Fig. S3). Furthermore, residual brain mass (and body size to a lesser extent) was significantly important within the phylogenetic models, but not in the non-phylogenetic models, suggesting that within a clade the species with the largest residual brain mass [@maklakov2011brains] and largest body size [@major2010museum] are most suited to surviving in urban areas. While some studies have found a lack of evidence that residual brain size is an important predictor in urban-tolerant species [@kark2007living; @evans2011makes], our results suggest that residual brain size is indeed heavily correlated with urban-tolerance [@moller2009successful; @maklakov2011brains; @moller2015brain]. This finding highlights the importance of investigating trait-urbanization relationships in a phylogenetically informed manner.

Certain aspects of the trait-urbanization relationship appear to be consistent, and thus generalizable, across the globe. For instance, our finding of the relative importance of annual fecundity is consistent with work in Europe [@croci2008does; @moller2009successful]. Urbanization appears to generally select against insectivorous species [@chace2006urban; @evans2011makes]. Although we did not find any strong support for our hypotheses on different nesting types, we found that species which tended to be more social were closely associated with urbanization, consistent with other findings [@coleman1994neophobia; @jokimaki1998distribution; @kark2007living]. While the aformentioned patterns may be generalizable, we also found patterns in our analysis which seem to contrast the global norm. For example, we found that urbanization selects against granivorous species, contrasting the general trend [@chace2006urban; @kark2007living; @evans2011makes], but this could be a result of relatively little supplementary feeding in Australian cities [cf., @fuller2008garden].

As illustrated above and in Appendix 1, many traits have been used to identify a species' adaptability to urban environments, but results have frequently differed among studies. As previously noted, a major shortcoming of previous studies has been the focus on discrete classifications of species based on their responses to urbanization. Our models assessed the applicability of traits in relation to one another (i.e., we did not run separate models or statistical tests for each specific trait as previously done; e.g., [@kark2007living]). The aforementioned differences in our results compared with others is likely due to increased sample size, as well as our use of a continuous measure of urbanization. Our generalizable analysis was made possible through the use of global broad-scale empirical datasets, which are rapidly expanding our knowledge of the impacts of urbanization on birds [e.g., @aronson2014global; @aronson2016hierarchical; @gutierrez2018citizen]. This is largely due to the growing use of citizen science data [@silvertown2009new] in combination with advances in remote sensing technology [@gorelick2017google]. We did not explicitly account for changes throughout the annual cycle in a species' association with urban environments [e.g., @la2014role; @lepczyk2017global; @katuwaleffect], but migratory status was considered as a predictor. We highlight that future studies, especially within migrant-dominated landscapes (e.g., Europe, North America), should account for changes in a species' response to urbanization throughout the full-annual cycle [@hostetler2015full; @marra2015call]. Further, future studies should look to expand this analysis outside the scope of Australia, and also investigate whether these patterns hold true at multiple spatial scales [@suarez2018scale] (i.e., does city size affect the patterns found here; [@sol2017urbanisation; @batary2018non]). Fully understanding the suite of traits which are associated with urban birds, and placing them in a theoretical landscape ecology context, will ultimately allow for a better understanding of how to best preserve and maintain avian biodiversity in urban areas [@evans2011makes]. We highlight that future studies investigating the trait-urbanization relationship should focus on broad-scale data to form generalizable patterns which rely on continuous measures for improved predictive performance. This work should be further extended to test whether our findings can accurately predict species which are commonly found in urban environments in other parts of the world, based on a known suite of traits associated with a given set of species.

We provide the largest study to date using a continuous urbanization index to assess ecological and life history traits which are associated with a species' adaptability to urban environments. Our results suggest that generalist species, with high clutch sizes, and large residual brain sizes, are better prepared to forage and survive in the ecologically novel aspects of urban environments [@shochat2006patterns]. In the context of our specific findings, Australia is a relatively young country with rapid urbanization currently underway [@coffee2016visualising] and fully understanding how birds respond to urban development is critical to inform future management. Urban avian biodiversity could be increased by improving the suitability of foraging for granivores and insectivores, especially important given the recent noted collapse of insects in other parts of the world [@hallmann2017more].

>>>>

---
title: "A continental-scale analysis finds generalists, with large clutch size and residual brain mass, are the most urban-tolerant of birds"

@@

# ABSTRACT
Understanding the suite of ecological and life history traits which influence a species' tolerance to urbanization is critical to understand which species are most likely to persist in our increasingly urbanizing world. Previous work investigating these relationships has found a wide array of contrasting patterns for a given trait's association with urbanization. In a continental-scale analysis, incorporating 477 species and > 5,000,000 bird observations, we provide a novel and scalable methodology which assesses the ecological and life history traits which most influence a species' adaptability to persist in urban environments. We elucidate generalized, phylogenetically controlled, patterns: bird species which are specialists, with high clutch size, and large residual brain size are among the most urban-tolerant bird species. Conversely, specialized feeding strategies (i.e., insectivores and granivores) are negatively associated with urbanization. Our results suggest that the enhancement and persistence of avian biodiversity in urban environments relies on maintaining and developing a suite of habitats to serve a range of life history strategies.

@@

In order to best predict species' responses to future urbanization changes, it is essential to understand what ecological and life history traits are most associated with birds that are either successful or unsuccessful in urban environments [e.g., @mcclure1989characterizes; @bonier2007urban; @moller2009successful;  @mcdonnell2015adaptation]. To date, a wide array of traits have been used to predict presence of birds in urban environments. For instance, degree of sociality [@mcclure1989characterizes; @coleman1994neophobia; @jokimaki1998distribution], nesting substrate [@kark2007living; @croci2008does; @conole2011functional], diet [@beissinger1982effects; @fuller2008garden; @major2010museum; @evans2011makes], distribution size [@jokimaki2000artificial; @chace2006urban; @croci2008does], migratory status [@friesen1995effects], niche breadth [@clergeau2006avifauna; @kark2007living; @evans2011makes], and fecundity [@croci2008does; @moller2009successful; @vaugoyeau2016interspecific], among other traits (see Appendix 1), have been hypothesized to influence a species' presence in urban environments. Rather than a specific trait that predicts a species' ability to be found in urban environments, it is most likely a combination of traits. For instance, the presence of urban birds in Jerusalem was predicted by traits including diet, degree of sociality, sedentariness, and preferred nesting sites [@kark2007living], while a regional analysis throughout France and Switzerland demonstrated that urban birds preferred forested habitats, were sedentary, omnivorous, high-elevation nesters, and widely distributed [@croci2008does]. Notably, a wide range of results using the aforementioned traits have been reported in the literature, often contrasting with one another. Indeed, residual brain size was found to be important by Moller and Erritze [-@moller2015brain] and Maklakov *et al.* [-@maklakov2011brains], but Evans *et al.* [-@evans2011makes] and Kark *et al.* [-@kark2007living] found that brain size had no effect on a species' response to urbanization. Similarly, contrasting results have been reported for annual fecundity [cf., @croci2008does; @moller2009successful; @evans2011makes] and niche breadth [cf., @kark2007living; @evans2011makes]. Forming generalizable patterns which associate ecological and life history traits with response to urbanization in birds, at multiple spatial scales [@suarez2018scale], is critical to help curb the impact urbanization has on avian biodiversity [@sol2017urbanisation].

The majority of previous studies assign species to discrete groups such as urban avoiders, adapters, or exploiters [e.g., @mckinney2002urbanization; @mckinney2006urbanization; @kark2007living; @mcdonnell2015adaptation] or even binary grouping based on presence/absence within urban environments [e.g., @bonier2007urban; @moller2009successful]. These types of discrete response variables severely limits the capacity to generalize results as it assumes species within a group respond to urbanization equally [@evans2011makes]. Moreover, the terminology (i.e., urban adapters, avoiders, and exploiters) is used differently among studies [cf., @kark2007living; @croci2008does], potentially affecting interpretation between scientists and the public [@mcdonnell2015adaptation; @fischer2015categorizing]. In a notable exception, Evans *et al.* [-@evans2011makes] advocates for a continuous measure of urbanization in which a species is assessed on its entire distribution in response to urban environments.

@@

The aim of this study is to determine whether there is a suite of life history traits that predicts avian tolerance to urbanization at a continental scale. For such an analysis to have generality it must (1) be comprehensive in faunal coverage, (2) account for phylogenetic constraints, (3), include a broad-range of traits, and (4) employ a continuous measure of response to urbanization. We conduct an analysis that meets these criteria, integrating a series of publicly available datasets to implement a novel continuous urbanization index in order to assess ecological and life history traits which most influence a bird's response to urbanization. Our first objective was to investigate the presence of a phylogenetic signal in our urbanization index, as previous studies have found a lack of phylogenetic signal in urbanization [@evans2011makes]. Our second objective was to develop models which were and were not phylogenetically controlled to assess species' traits and their association with urbanization, using a suite of continuous and categorical predictor variables. We hypothesized that a distinct suite of particular traits would be positively associated with urbanization (i.e., urbanization selects for) while a separate suite of traits would be negatively associated with urbanization (i.e., urbanization selects against).

@@

We weighted the linear regression models [@solon2015weighting] using a source of error associated with our urbanization indices. This error was the  proportion of the number of total observations of a species divided by the number of unique localities a species was observed from. This gave more weight in the model-fitting procedure to those species that had more observations in more unique locations. This was done based on the assumption that those species which had the most observations would have the most reliable urbanization indices. This approach also addresses biases in sampling of eBird data in which some birds can have a high number of observations, but from known roosting sites [e.g., @callaghan2018avian].

@@

# DISCUSSION
Niche breadth, as measured by different types of generalism, was a significantly important life-history characteristic associated with the most urban-tolerant species in our analyses. This result was robust across non-phylogenetic and phylogenetic models, model-averaging approaches, and across phylogenetic uncertainty (Fig. S3). Conversely, specialized feeding strategies were negatively associated with urbanization: granivorous and insectivorous specialists were much less likely to be found in urban environments. These results suggest that the process of urbanization is, in general, operating to exclude specialist species [@sorace2009distribution]. Given specialist species are known to have strong negative responses to landscape fragmentation and disturbance [@devictor2008distribution], sustainable urban development with a goal of maintaining biological diversity in urban areas should focus on providing intact vegetation [@sushinsky2013should] connected by corridors [@litteral2017role]. A species-specific approach, which assesses particular resources necessary for specialist species, for example focusing on foraging [@shochat2004linking] and nesting [@fernandez2001habitat] requirements, is likely necessary. We recommend that a focal point of avian biodiversity conservation should be on increasing insect abundance and diversity in urban environments [@baldock2015uk], especially important given the recent noted collapse of insects in other parts of the world [@hallmann2017more].

The non-random distribution of the urbanization index across the phylogeny (Table 1) is consistent with the observation of reduced phylogenetic diversity in urban areas [@la2018phylogenetic], suggesting that particular phylogenetic constraints allow certain species to remain tolerant of urban environments [@morelli2016evidence; @ibanez2017global; @sol2017urbanisation]. This finding of a strong phylogenetic signal contrasts with previous studies [@evans2011makes], but this is likely a result of increased statistical power from our larger sample size: 477 species in this study compared with 55 species in Evans *et al.* [-@evans2011makes]. Unsurprisingly, our larger sample size found a non-random pattern in urbanization across the bird phylogeny and we further demonstrated the importance of treating species as non-independent by considering phylogeny in the context of the linear models. For example, some predictor variables became less important in the phylogenetic models compared with the non-phylogenetic models, including some feeding types (i.e., insect specialists) and habitat preferences (i.e., agricultural). For these variables, it is likely that there may be particular clades which include a number of species that are associated with both that particular trait and degree of urbanization [see @felsenstein1985confidence]. Conversely, some of the most important predictor variables (e.g., clutch size and feeding habitat generalism; Figures 4 and 5) showed a stronger relationship (i.e., increased parameter estimates) with urbanization when phylogeny was accounted for, and this affect was robust relative to a consideration of phylogenetic uncertainty (Fig. S3). Furthermore, residual brain mass (and body size to a lesser extent) was significantly important within the phylogenetic models, but not in the non-phylogenetic models, suggesting that within a clade the species with the largest residual brain mass [@maklakov2011brains] and largest body size [@major2010museum] are most suited to surviving in urban areas. While some studies have found a lack of evidence that residual brain size is an important predictor in urban-tolerant species [@kark2007living; @evans2011makes], our results suggest that residual brain size is indeed heavily correlated with urban-tolerance [@moller2009successful; @maklakov2011brains; @moller2015brain]. This finding highlights the importance of investigating trait-urbanization relationships in a phylogenetically informed manner.

Certain aspects of the trait-urbanization relationship appear to be consistent, and thus generalizable, across the globe. For instance, our finding of the relative importance of annual fecundity is consistent with work in Europe [@croci2008does; @moller2009successful]. Urbanization appears to generally select against insectivorous species [@chace2006urban; @evans2011makes]. Although we did not find any strong support for our hypotheses on different nesting types, we found that species which tended to be more social were closely associated with urbanization, consistent with other findings [@coleman1994neophobia; @jokimaki1998distribution; @kark2007living]. While the aforementioned patterns may be generalizable, we also found patterns in our analysis which seem to contrast the global norm. For example, we found that urbanization selects against granivorous species, contrasting with the general trend [@chace2006urban; @kark2007living; @evans2011makes], but this could be a result of relatively little supplementary feeding in Australian cities [cf., @fuller2008garden].

As illustrated above and in Appendix 1, many traits have been used to identify a species' adaptability to urban environments, but results have frequently differed among studies. As previously noted, a major shortcoming of previous studies has been the focus on discrete classifications of species based on their responses to urbanization. Our models assessed the applicability of traits in relation to one another (i.e., we did not run separate models or statistical tests for each specific trait as previously done; e.g., [@kark2007living]). The aforementioned differences in our results compared with others is likely due to increased sample size, as well as our use of a continuous measure of urbanization. Our broad-ranging analysis was made possible through the integration of broad-scale empirical datasets, which are rapidly expanding our knowledge of the impacts of urbanization on birds [e.g., @aronson2014global; @aronson2016hierarchical; @gutierrez2018citizen]. This is largely due to the growing use of citizen science data [@silvertown2009new] in combination with advances in remote sensing technology [@gorelick2017google]. We did not explicitly account for changes throughout the annual cycle in a species' association with urban environments [e.g., @la2014role; @lepczyk2017global; @katuwaleffect], but migratory status was considered as a predictor. We emphasize that future studies, especially within migrant-dominated systems (e.g., Europe, North America), should account for changes in a species' response to urbanization throughout the full-annual cycle [@hostetler2015full; @marra2015call]. Further, future studies should look to expand this analysis to other continents, and also investigate whether these patterns hold true at multiple spatial scales [@suarez2018scale] (i.e., does city size affect the patterns found here; [@sol2017urbanisation; @batary2018non]). Fully understanding the suite of traits which are associated with urban-tolerant species, and placing them in a theoretical landscape ecology context will ultimately allow for a better understanding of how to best preserve and maintain avian biodiversity in urban environments [@evans2011makes]. We highlight that future studies investigating the trait-urbanization relationship should focus on broad-scale data to form generalizable patterns which rely on continuous measures for improved predictive performance. This work should be further extended to test whether our findings can accurately predict species which are commonly found in urban environments in other parts of the world, based on a known suite of traits associated with a given set of species.

We provide the largest study to date using a continuous urbanization index to assess ecological and life history traits which are associated with a species' adaptability to urban environments. Our results suggest that generalist species, with high clutch sizes and large residual brain sizes, are best prepared to forage and survive in the ecologically novel aspects of urban environments [@shochat2006patterns]. In the context of our specific findings, Australia is a relatively young country with rapid urbanization currently underway [@coffee2016visualising] and fully understanding how birds respond to urban development is critical to inform future management. Urban avian biodiversity could be increased by emphasizing the critical importance of intact vegetation [@sushinsky2013should] for specialist species [@devictor2008distribution], and concomitantly improving the suitability of foraging for granivores and insectivores.
